\chapter{Preface}
\ifdefstring{\releasename}{Frozen}{
    {\bf
    This specification is Frozen.

    Change is extremely unlikely. A high threshold will be used, and a change
    will only occur because of some truly critical issue being identified during
    the public review cycle. Any other desired or needed changes can be the
    subject of a follow-on new extension.}
}{}

\ifdefstring{\releasename}{STABLE}{
    {\bf This stable specification might change before being accepted as
    standard. We're trying to only make backwards-compatible changes, but
    sometimes it's unavoidable to change a corner case. If you implement
    this document, please keep an eye on
    \href{https://github.com/riscv/riscv-debug-spec}{the github repository} for
    changes that might affect your implementation.}
}{}
\ifdefstring{\releasename}{DRAFT}{
    {\bf Warning! This draft specification will change before being accepted as
    standard, so implementations made to this draft specification will likely not
    conform to the future standard.}
}{}

Contributors to all versions of the spec in alphabetical order (please contact
editors to suggest corrections):
Bruce Ableidinger,
Krste Asanovi\'{c},
Peter Ashenden,
Allen Baum,
Mark Beal,
Alex Bradbury,
Chuanhua Chang,
Yen Hao Chen,
Zhong-Ho Chen,
Monte Dalrymple,
Paul Donahue,
Vyacheslav Dyachenko,
Ernie Edgar,
Peter Egold,
Marc Gauthier,
Markus Goehrle,
Robert Golla,
John Hauser,
Richard Herveille,
Yung-ching Hsiao,
Po-wei Huang,
Scott Johnson,
L. J. Madar,
Grigorios Magklis,
Daniel Mangum,
Alexis Marquet,
Jan Matyas,
Kai Meinhard,
Jean-Luc Nagel,
Aram Nahidipour,
Rishiyur Nikhil,
Gajinder Panesar,
Deepak Panwar,
Antony Pavlov,
Klaus Kruse Pedersen,
Ken Pettit,
Darius Rad,
Joe Rahmeh,
Josh Scheid,
Vedvyas Shanbhogue,
Gavin Stark,
Ben Staveley,
Wesley Terpstra,
Tommy Thorn,
Megan Wachs,
Jan-Willem van de Waerdt,
Philipp Wagner,
Stefan Wallentowitz,
Ray Van De Walker,
Andrew Waterman,
Thomas Wicki,
Andy Wright,
Bryan Wyatt,
and Florian Zaruba.
